---
title: "Regular Expression"
editor: visual
---

Regular expression is a pain to programmers who are only occasionally users. You are most likely forgetting the rules after 3 months not using it. This document helps to better understand regular expression so that you can easily pick it up after not using it even for one year.

## How regex engine matches patterns

### Basic operation

A match start from the leftmost of an input string and moves from left to right until the first match is found.

In the example below, the match starts from `^`, the line beginning of input string `x`. As there is no match for the first letter `c` of pattern `cat`, the regex engine moves to `A`, the first letter of string `x`. Still it does match `c` so the engine keeps move left to `c` in `black` of string `x`. Now `c` matches the first letter of `c` in the pattern `cat`, the engine checks second letter in the patter, `a`. It does not match `k` after `c` in string `x` so this attempt fails. The engine now moves to `k` in the input string and tries to match the pattern `cat` all over again. The engine repeats this process until find the first match, `cat` in `catfish`, and report the match as the results. By default, the engine stops searching for additional matches.

```{r}
x <- "A black catfish for an orange cat"
stringr::str_extract(x, "cat")
```

### Backtracking in greedy match

When the search pattern contains unbounded quantifiers such as `*`, `+`, and `{3, }`, the engine tries to match as many as possible. It then tracks back to match the pattern.

In the example below, the `.+` matches the whole string `x`. It then tries to match `cat` and fails. The engine now track back to a previous good match for `.+`, which is `A black catfish for an orange ca`. Still there is no match for `cat`. So the engine continuous tracks back to `A black catfish for an orange c`. This backtracking is repeated until a match is found for `cat`.

```{r}
x <- "A black catfish for an orange cat"
stringr::str_extract(x, ".+cat")
```

### Catastrophic backtracking

Catastrophic backtracking may happen when there is no match. A regex engine reports no matching only after exhausting all possibilities. So it may take a very long time to finish. In the example below, it take 1 second to finish when x has 24 `a`s, and 8 seconds when 27 `a`s.

```{r}
system.time({
    x <- "aaaaaaaaaaaaaaaaaaaaaaaa"   # 24 a
    stringr::str_extract(x, "(a+a+)+y")
})

system.time({
    x <- "aaaaaaaaaaaaaaaaaaaaaaaaaaa"   # 27 a
    stringr::str_extract(x, "(a+a+)+y")
})
```

### Lazy match

A greedy match pattern is turned into a lazy match pattern if a `?` is appended to it, like `*?`, `+?`, and `{3, }?`. The engine tries to match as few as possible.

In the example below, the engine starts at `A` and tries to make a match for `.*?`. As the `.*` requires as few as 0 of any character, so the engine first match it to empty. Because of lazy match, the engine is now look to match `c` in the pattern to `A` in the string. Of course if fails. Now the engine expands `.*?` to include `A` and check `c` against space but still fails. So the engine expands `.*?` to include `A b` and repeat the process until `cat` find a match in the string. The first match `A black cat` is the return of the match.

```{r}
x <- "A black catfish for an orange cat"
stringr::str_extract(x, ".*?cat")
```

## Capture groups

### What is capture groups

Capture groups are used when we want the regex engine to return sub patterns in addition to the overall matching pattern. In regex pattern, a capturing group lives in a pair of round parentheses `(pattern)`.

The following example has two capture groups and three strings are returned, The first string is `123-4567890`, which is the overall matching pattern and is called group 0. The second returned string is `123`, which matches capturing group 1 `(\\d{3})`. The third is `4567890`, which matches group 2 `(\\d{7})`.

```{r}
x <- c("123-4567890", "401-7896787")
matches <- stringr::str_match(x, "(\\d{3})-(\\d{7})")
matches
```

```{r}
area_code <- matches[, 2]
area_code
```

### Named capture groups

We can give each or some capture groups a name instead of just group 2 or 3.

```{r}
x <- "123-4567890"
stringr::str_match(x, "(?<areaCode>\\d{3})-(?<number>\\d{7})")  # area_code not supported
```

```{r}
# area code
stringr::str_match(x, "(?<areaCode>\\d{3})-(?<number>\\d{7})")[, "areaCode"]
```

### Back referencing

The captured groups 1 and 2 can be referenced as `\\1` and `\\2` in subsequent matching patterns that point to the exact captured contents .Here are a few examples:

In the example below we switch the order of first name and last name using back references in a search and replacement.

```{r}
x <- "Hanks, Tom"
stringr::str_replace(x, "(\\w+),\\s(\\w+)", "\\2 \\1")
```

Another use case of back referencing is to find duplicated words.

```{r}
x <- "what if if I am am wrong ong"
stringr::str_extract_all(x, "\\b(\\w+)\\s\\1\\b")  # need word boundary 
```

And we can delete duplicated words

```{r}
x <- "what snif if if I am am wrong"
stringr::str_replace_all(x, "\\b(\\w+)\\s\\1\\b", "\\1")
```

### Nested capture groups

A capture group can contain one or more capture groups inside it. They are ordered by the position of their open parenthesis in back reference. In the example below, `((\d{3})-(\\d{7}))` is `\\1`, `(\\d{3})` is `\\2`, and `(\\d{7})\` is`\\3`.

```{r}
# extract true number and area code and number
x <- "Fake number is 123-4567890 and true number is 401-1234567"
stringr::str_match(x, "true.+((\\d{3})-(\\d{7}))")
```
